---
title: "Final Assignment DCM - Young Migrants and Degree Choice"
author: "Marc Sparhuber"
format: pdf
editor: source
execute:
  warning: false
  echo: false
toc: true
header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---

\newpage 

```{r}
library(tidyverse)
library(tidymodels)
library(survival)
library(kableExtra)
library(gtsummary)
library(haven)

og_data <- read_dta(here::here("DCM","FINALDCM", "acs_2015.dta"))

og_data$degfield <- labelled::to_factor(og_data$degfield)
og_data$degfieldd <- labelled::to_factor(og_data$degfieldd)
og_data$degfield2 <- labelled::to_factor(og_data$degfield2)
og_data$degfield2d <- labelled::to_factor(og_data$degfield2d)
og_data$empstat <- labelled::to_factor(og_data$empstat)
og_data$rachsing <- labelled::to_factor(og_data$rachsing)
og_data$stateicp <- labelled::to_factor(og_data$stateicp)
og_data$met2013 <- labelled::to_factor(og_data$met2013)
og_data$bpld <- labelled::to_factor(og_data$bpld)
og_data$bpl <- labelled::to_factor(og_data$bpl)
og_data$occ <- labelled::to_factor(og_data$occ)
og_data$citizen <- labelled::to_factor(og_data$citizen)
og_data$sex <- labelled::to_factor(og_data$sex)
og_data$educ <- labelled::to_factor(og_data$educ)
```

## Introduction

> With migration expected to rise drastically in the coming decades, issues of integration continously rise in relevance [SOURCE]. One aspect of migration is educational attainment and more specifically choice of degree field in tertiary education, as getting a degree serves as a vehicle for upward social mobility [SOURCE]. Especially for individuals who hae migrated at a young age, who have yet to make their decisions regarding their education, different reasons for degree choice may be paramount compared to non-migrant children of their age cohort. Thus, the central research question of this research paper is whether individuals who have migrated at a young age differ in their choice of tertiary education to non-migrants. Connected to this are the questions of whether their access to education is the same as that of their non-migrant peers and what hopes and attributions are connected to certain degree fields.

> To answer these questions, binary, multinomial and conditional logistic regressions are employed. The analysis is based on data from the American Community Survey 2015, which is a demographic survey conducted yearly by the United States Census Bureau. It contains individual and household-level information about demographic, educational, and financial information, among other fields. Based on some key variables ... [TEASER OF RESULTS HERE].

## Background

> [Find something about
- educational assimilation
- degree choice
- utility as reason for choice
]

> Something about the specific case: in the US - costs associated with tertiary education high -> might be more utility to get an apprenticeship but there may be greater expectations attached, with people believing in the myth of the american dream? -> I.E., college as the main road to success

> It is expected that though there may be fewer young migrants choosing to pursue tertiary education due to the associated costs compared to their non-migrant counterparts (binary), those who do are driven by strong sense of utility in their choice of studies, therefore choosing fields commonly associated with higher prestige and better financial stability (multi). Additionally, the attributed outcomes associated with degree choice are expected to primarily be related to financial stability and prestige (clog).

> This analysis has a breadth of potential confounders and mediators to take into account, many of which are expected to interact with one another and not all of which are included in the data. Sex is expected to serve as a strong confounder, as several fields of study are strongly gender segregated [IS THAT HOW YOU SAY IT?]. This effect is expected to differ, however, based on the socialization at home, where variables such as religion, country of birth as well as parent related variables may show interactions with sex and have their own confounding effects. Citizenship status may influence whether college is attended at all, as undocumented students may choose to not attend college due either not thinking that they are allowed to do so or because their funding options by the state are reduced [https://counselors.collegeboard.org/financial-aid/undocumented-students]. Additionally, degree choices may, however, be greatly influenced by the age at which an individual has migrated to the US country. Due to this, some of the following models take this into account.



```{r desctable}

data_descriptives <- og_data

# Deal with Nas and empty levels
data_descriptives$degfield <- fct_recode(data_descriptives$degfield, NULL = "N/A")
data_descriptives$citizen <- fct_recode(data_descriptives$citizen, "Citizen" = "N/A")
data_descriptives$empstat <- fct_recode(data_descriptives$empstat, NULL = "N/A")

data_descriptives$citizen <- droplevels(data_descriptives$citizen)
data_descriptives$empstat <- droplevels(data_descriptives$empstat)
data_descriptives$educ <- droplevels(data_descriptives$educ)
data_descriptives$sex <- droplevels(data_descriptives$sex)

 
data_descriptives <- data_descriptives |> 
  select(sex,age,stateicp,met2013,birthyr,bpl,citizen,yrimmig,yrsusa1,rachsing,educd,degfield,empstat,uhrswork,inctot,incearn,poverty,educ) |>
  mutate(
   age_at_immigration = if_else(yrimmig != 0, yrimmig - (2015-age), NA_integer_),
   young_migrant = if_else(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 19, "YM", "Not YM"),
   young_migrant = as_factor(young_migrant),
   young_migrant_detailled = case_when(young_migrant == "YM" & age_at_immigration < 10 ~ "Migrated as Child (younger than 10 years old)",
                             young_migrant == "YM" & age_at_immigration >= 10 & age_at_immigration <= 19 ~ "Migrated as Adolescent (between 10 and 19 years old)",
                             young_migrant == "Not YM" ~ "Not YM"),
   young_migrant_detailled = fct_relevel(young_migrant_detailled, "Not YM"),
   `Migration Status` = case_when(
     !is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 19 ~ "Migrated Young (younger than 20)",
     !is.na(age_at_immigration) & age_at_immigration > 19 ~ "Migrated as Adult (20 and older)",
     TRUE ~ "Did not Migrate"
   ),
   educ = fct_collapse(educ,
                       other_level = "Did not Finish High School",
                       `Finished High School` = c("Grade 12"),
                       `Some College` = c("1 year of college", "2 years of college"),
                       `Bachelor's Degree` = c("4 years of college"),
                       `Master's Degree or Doctorate` = c("5+ years of college")),
   educ = fct_relevel(educ, "Did not Finish High School")
   ) |> 
  rename(`Race` = rachsing) |> 
  filter(birthyr >= 1960 & age >= 24 & inctot < 9999998)


data_descriptives |>
  select(`Migration Status`,`Age (in years)` = age, `Sex` =  sex, `Age (in years)` = age, `Citizenship Status` = citizen, `Race`, `Educational Attainment` = educ, `Employment Status` = empstat, uhrswork, inctot, incearn, `Poverty Status` = poverty) |> 
  tbl_summary(by = `Migration Status`, type = list(`Age (in years)` ~ "continuous",
                                                   uhrswork ~ "continuous",
                                                   inctot ~ "continuous",
                                                   `Poverty Status` ~ "continuous")) |>
  modify_caption("Educational and Economical Indicators: Descriptives Divided by Migration Status") |>
  as_kable_extra(booktabs = TRUE, escape = FALSE, linesep = "") |>
  footnote(footnote_as_chunk = TRUE,
           threeparttable = TRUE, 
           general = "Shows only those respondents who were born after 1960 and at least 24 years old in 2015. Poverty Status is calculated taking into account household size, household member's age, age of the householder and total family income: Values over 100 indicate being above the poverty threshold. Data from the American Community Survey 2015 (collected 2015).")
```

## Data

> The data for the 2015 American Community Survey was gathered over the course of the year 2015, containing more than 3 million individuals and is representative for the inhabitants of the Unites States of America.

Describe how you generated your analytic sample from the data
> Of these, such and such data are used -> birthyear, age, etc. (slightly varies across analyses due to type of regression, for example looking only at employed individuals for multi).

Explain how you treated missing values.
> Individuals containing missing values in any of the variables of interest are dropped from the analysis.

Clearly identify the dependent variable(s).
> Many, list the final ones used here
> how did I group degrees? (only for multi)

Identify and explain the operationalization of key independent variables.
> young_migrant, also mention country collapse

Produce and discuss descriptive statistics.


```{r clogit}
# What are the alternative specific attributes (i.e., the outcomes associated with degree choice) that are considered desirable by choice makers (and their parents)?

data_clog <- og_data |> select(sex,age,stateicp,met2013,birthyr,bpl,citizen,yrimmig,
                          yrsusa1,rachsing,educd,degfield,empstat,uhrswork,inctot,incearn,poverty)

#drop na level in degree so it doesn't get annoying later
data_clog$degfield <- fct_recode(data_clog$degfield, NULL = "N/A")
data_clog$citizen <- fct_recode(data_clog$citizen, "Citizen" = "N/A")
data_clog$empstat <- fct_recode(data_clog$empstat, NULL = "N/A")

# reduce to people who have done a bachelors at least - they will be the observed group
data_clog <- data_clog |> filter(educd >= 101 & educd != 999)

# make new variable to get age at migration and take a look at my group of interest
data_clog <- data_clog |> mutate(
   age_at_immigration = if_else(yrimmig != 0, yrimmig - (2015-age), NA_integer_),
   young_migrant = if_else(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 19, "YM", "Not YM"),
   young_migrant = as_factor(young_migrant),
   young_migrant_detailled = case_when(young_migrant == "YM" & age_at_immigration < 10 ~ "Migrated as Child (younger than 10 years old)",
                             young_migrant == "YM" & age_at_immigration >= 10 & age_at_immigration <= 19 ~ "Migrated as Adolescent (between 10 and 19 years old)",
                             young_migrant == "Not YM" ~ "Not YM"),
   young_migrant_detailled = fct_relevel(young_migrant_detailled, "Not YM")
   ) |> 
  filter(birthyr >= 1960)


data_clog$empstat <- droplevels(data_clog$empstat)


# make a little summarize table
interesting_data_smile <- data_clog |>
  group_by(degfield) |> 
  filter(n()>=1000) |> 
  summarise(
    n_degree_holders = n(),
    med_age = median(age),
    n_young_migrants = sum(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21),
    p_young_migrants = sum(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21)/n(),
    n_employed = sum(empstat == "Employed"),
    p_employed = sum(empstat == "Employed")/n(),
    n_ym_employed = sum(empstat == "Employed" & !is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21),
    p_ym_employed_within = sum(empstat == "Employed" & !is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21)/n_young_migrants,
    p_ym_employed_general = sum(empstat == "Employed" & !is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21)/n(),
    mean_weekly_hours = round(mean(uhrswork),2),
    mean_tot_income = mean(inctot),
    mean_earned_income = mean(incearn),
    mean_poverty = round(mean(poverty),2)
    )


#so there are no too small categories blah
interesting_data_smile$degfield <- droplevels(interesting_data_smile$degfield)

# id part!
set.seed(1)
sample_degree_holders <- data_clog |> filter(degfield %in% interesting_data_smile$degfield)
sample_degree_holders$degfield <- fct_drop(sample_degree_holders$degfield)

sample_degree_holders <- sample_degree_holders |> 
group_by(young_migrant) |> 
  slice_sample(n = 5000) |> ungroup() |> mutate(ID = row_number())

size <- 5
alts <- interesting_data_smile$degfield

sample_degree_holders |> 
  ungroup() |> 
  select(ID) |>  
  mutate(alt = replicate(n(), sample(alts, size = size, replace = FALSE), simplify = FALSE)) |> 
  unnest(alt) |>
  anti_join(sample_degree_holders, by = join_by(ID, alt == degfield)) |> 
  group_by(ID) |> slice(1:(size-1)) ->
  degree_choice


# reduce variables in degree table for conditional logit
interesting_data_smile2 <- interesting_data_smile 
#|> select(-c(starts_with("n_"),starts_with("p_"),med_age))


sample_degree_holders |> 
  select(ID, alt = degfield) |>
  bind_rows(degree_choice) |> 
  group_by(ID) |> 
  arrange(ID, alt) |> 
  left_join(sample_degree_holders, by = join_by(ID)) |> 
  left_join(interesting_data_smile2, by = join_by(alt == degfield)) |> 
  mutate(choice = degfield == alt,
         choice = as.numeric(choice)) |> 
  select(-c(yrimmig,yrsusa1,met2013,educd,degfield,stateicp,bpl,age_at_immigration)) -> degree_choice2

# Sanity check good
sum(degree_choice2$choice) == nrow(sample_degree_holders)
sum(degree_choice2$choice)*size == nrow(degree_choice2)

library(survival)

# OKAY SO BIG REVELATION:
# don't include alt duh
# only look at what makes the choice, not the choice itself
# these models often don't work cause the multicol from including the alts itself

m_clog_inter <- clogit(choice ~ mean_weekly_hours:young_migrant + mean_poverty:young_migrant + mean_earned_income:young_migrant + mean_tot_income:young_migrant + p_employed:young_migrant + strata(ID), data = degree_choice2)
tidy(m_clog_inter) |> View()

tbl_regression(m_clog_inter, exponentiate = TRUE) |>
  modify_header(list(label = "**Covariate**",
                     estimate = "**Odds**")) |>
  modify_footnote(update = list(estimate ~ NA),
                  abbreviation = TRUE)
```



```{r multinom}
# which degrees do migrants who enter the country young enter into more and what are their attributes?

data_mult <- og_data |> select(sex,age,stateicp,met2013,birthyr,bpl,citizen,yrimmig,
                          yrsusa1,rachsing,educd,degfield,empstat,uhrswork,inctot,incearn,poverty)


#drop na level in degree so it doesn't get annoying later
data_mult$degfield <- fct_recode(data_mult$degfield, NULL = "N/A")
data_mult$citizen <- fct_recode(data_mult$citizen, "Citizen" = "N/A")
data_mult$empstat <- fct_recode(data_mult$empstat, NULL = "N/A")


# reduce to people who have done a bachelors at least - they will be the observed group
data_mult <- data_mult |> filter(educd >= 101 & educd != 999)


# make new variable to get age at migration and take a look at my group of interest
data_mult <- data_mult |> mutate(
   age_at_immigration = if_else(yrimmig != 0, yrimmig - (2015-age), NA_integer_),
   young_migrant = if_else(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 19, "YM", "Not YM"),
   young_migrant = as_factor(young_migrant),
   young_migrant_detailled = case_when(young_migrant == "YM" & age_at_immigration < 10 ~ "Migrated as Child (younger than 10 years old)",
                             young_migrant == "YM" & age_at_immigration >= 10 & age_at_immigration <= 19 ~ "Migrated as Adolescent (between 10 and 19 years old)",
                             young_migrant == "Not YM" ~ "Not YM"),
   young_migrant_detailled = fct_relevel(young_migrant_detailled, "Not YM")
   ) |> 
  filter(birthyr >= 1960)



#so there are no too small categories blah
data_mult$degfield <- droplevels(data_mult$degfield)
data_mult$bpl <- droplevels(data_mult$bpl)

# make broader categories for degfield
data_mult <- data_mult |> mutate(
  degfield_collapsed = fct_collapse(degfield,
    Engineering = c("Architecture", "Communication Technologies", "Engineering", "Engineering Technologies", "Military Technologies", "Nuclear, Industrial Radiology, and Biological Technologies", "Construction Services", "Electrical and Mechanic Repairs and Technologies", "Transportation Sciences and Technologies"),
    Sciences = c("Agriculture", "Environment and Natural Resources", "Computer and Information Sciences", "Biology and Life Sciences", "Physical Sciences", "Mathematics and Statistics"),
    Arts = c("Area, Ethnic, and Civilization Studies", "Cosmetology Services and Culinary Arts", "Linguistics and Foreign Languages", "English Language, Literature, and Composition", "Liberal Arts and Humanities", "Library Science", "Physical Fitness, Parks, Recreation, and Leisure", "Philosophy and Religious Studies", "Fine Arts", "History"),
    `Social Sciences` = c("Communications", "Family and Consumer Sciences", "Psychology", "Public Affairs, Policy, and Social Work", "Social Sciences"),
    Medicine = c("Medical and Health Sciences and Services"),
    Business = c("Business"),
    Education = c("Education Administration and Teaching"),
    other_level = "Other"
  ),
  degfield_collapsed = fct_relevel(degfield_collapsed, "Business"),
  sex = as_factor(sex),
  bpl_collapsed = fct_collapse(bpl,
                     other_level = "North America",
                     LatAm = c("Mexico","Central America","Cuba","Americas, n.s.","West Indies","SOUTH AMERICA"),
                     `Western Europe & Scandinavia` = c("Denmark", "Finland", "Iceland", "Norway", "Sweden", "England", "Scotland", "United Kingdom, ns", "Ireland", "Belgium", "France", "Netherlands", "Switzerland", "Greece", "Italy","Portugal", "Spain", "Austria", "Germany"),
                     `Eastern Europe` = c("Albania", "Bulgaria", "Czechoslovakia", "Hungary", "Poland", "Romania", "Yugoslavia", "Latvia", "Lithuania", "Other USSR/Russia", "Europe, ns", "Cyprus"),
                     `East Asia` = c("China", "Japan", "Korea"),
                     `South & South-East Asia` = c("India" ,"Cambodia (Kampuchea)", "Indonesia", "Laos", "Malaysia","Philippines","Singapore","Thailand","Vietnam", "Nepal", "Asia, nec/ns"),
                     `Middle East` = c("Afghanistan", "Iran","Iraq","Israel/Palestine", "Jordan","Kuwait","Lebanon","Saudi Arabia","Syria","Turkey","United Arab Emirates", "Yemen Arab Republic (North)"),
                     Africa = c("AFRICA"),
                     Oceania = c("Australia and New Zealand", "Pacific Islands")
  ),
  bpl_collapsed = fct_relevel(bpl_collapsed, "North America")
)

data_mult$sex <- droplevels(data_mult$sex)


data_mult |> select(sex, age, citizen, rachsing, degfield, degfield_collapsed, empstat, uhrswork, inctot, incearn, poverty, young_migrant, bpl_collapsed,birthyr, young_migrant_detailled) |> 
  filter(empstat == "Employed" & poverty != 0) -> data_mult

data_mult <- data_mult |> group_by(young_migrant) |> slice_sample(n = 20000) |> ungroup()

library(nnet)
library(broom)
library(modelr)

# YOUNG MIGRANT NEEDS TO BE SPLIT INTO DIFFERENT VALUES DEPENDING ON AGE -> THE YOUNGER THE MORE ADJUSTED?
multi <- multinom(degfield_collapsed ~ I(sex:young_migrant) + young_migrant_detailled + bpl_collapsed, data = data_mult, trace = F, model = T, na.action = na.exclude)
#tidy(multi) |> View()

modelsummary::modelsummary(multi, shape =  term ~ model + response,
             fmt = 2,
             statistic = 'conf.int',
             title = "title",
             exponentiate = TRUE,
             stars = TRUE)


# TODO:
# ADD GOF TABLE HERE
# ADD PREDICTION GRAPH HERE
```


```{r binary}
# here just compare the YM with other people - how are they unique? do they earn more or less? are they more poor?
# I.E., regardless of expectation - what does it look like for YM?

data_binary <- og_data |> select(sex,age,stateicp,met2013,birthyr,bpl,citizen,yrimmig,
                          yrsusa1,rachsing,educd,degfield,empstat,uhrswork,inctot,incearn,poverty)

#drop na level in degree so it doesn't get annoying later
data_binary$degfield <- fct_recode(data_binary$degfield, NULL = "N/A")
data_binary$citizen <- fct_recode(data_binary$citizen, "Citizen" = "N/A")
data_binary$empstat <- fct_recode(data_binary$empstat, NULL = "N/A")

# reduce to people who have done a bachelors at least - they will be the observed group
data_binary <- data_binary |> filter(educd >= 101 & educd != 999)

# make new variable to get age at migration and take a look at my group of interest
data_binary <- data_binary |> mutate(
   age_at_immigration = if_else(yrimmig != 0, yrimmig - (2015-age), NA_integer_),
   young_migrant = if_else(!is.na(age_at_immigration) & age_at_immigration >= 0 & age_at_immigration <= 21, "YM", "Not YM"),
   young_migrant = as_factor(young_migrant)
   ) |> 
  filter(birthyr >= 1960)

#TODO:
# DO MODEL
# ADD GOF TO MODELSUMMARY HERE
```




```{r the bin}
# degree_choice3 <- degree_choice2 |> select(ID, choice, rachsing, alt, young_migrant)
# 
# 
# # why does one work an the other doesn't?
# # https://stats.stackexchange.com/questions/377997/na-coefficients-produced-by-clogit-model
# this_works <- clogit(choice ~ mean_weekly_hours + mean_poverty+ mean_earned_income + alt + strata(ID), data = degree_choice2)
# tidy(this_works) |> View()
# 
# this_doesnt_work <- clogit(choice ~ I(alt:young_migrant) + mean_weekly_hours + mean_poverty + mean_earned_income + strata(ID), data = degree_choice2)
# tidy(this_doesnt_work) |> View()
# 
# 
# 
# degree_choice2 %>%
#   group_by(alt) %>%
#   summarize(Num_Unique_Values = n_distinct(uhrswork)) |> View()
# 
# 
# 
# # only person characteristics -> does not work
# person_only <- clogit(choice ~ I(as.numeric(uhrswork)) + strata(ID), data = degree_choice2)
# tidy(person_only) |> View()
# 
# 
# 
# # filtered stuff
# ym <- clogit(choice ~ mean_weekly_hours + mean_poverty+ mean_earned_income + alt + strata(ID) ,
#        data = filter(degree_choice2, young_migrant == "YM"))
# tidy(ym) |> View()
# 
# t1 <- tbl_regression(ym, exponentiate = TRUE) |> 
#   modify_header(list(label="**Covariate**",
#                      estimate="**Odds**")) |>
#   modify_footnote(update = list(estimate ~ NA), 
#                   abbreviation=TRUE)
# t1
# 
# nym <- clogit(choice ~  mean_weekly_hours + mean_poverty+ mean_earned_income + alt + strata(ID) ,
#        data = filter(degree_choice2, young_migrant == "Not YM"))
# tidy(nym) |> View()
# 
# t2 <- tbl_regression(nym, exponentiate = TRUE) |> 
#   modify_header(list(label="**Covariate**",
#                      estimate="**Odds**")) |>
#   modify_footnote(update = list(estimate ~ NA), 
#                   abbreviation=TRUE)
# t2


not_ym_m <- clogit(choice ~ mean_weekly_hours + mean_poverty + mean_earned_income + mean_tot_income + p_employed + strata(ID), data = filter(degree_choice2, young_migrant == "Not YM"))
tidy(not_ym_m) |> View()

t_nym <- tbl_regression(not_ym_m, exponentiate = TRUE) |>
  modify_header(list(label="**Covariate**",
                     estimate="**Odds**")) |>
  modify_footnote(update = list(estimate ~ NA),
                  abbreviation=TRUE)
t_nym


ym_m <- clogit(choice ~ mean_weekly_hours + mean_poverty + mean_earned_income + mean_tot_income + p_employed + strata(ID), data = filter(degree_choice2, young_migrant == "YM"))
tidy(not_ym_m) |> View()

t_ym <- tbl_regression(ym_m, exponentiate = TRUE) |>
  modify_header(list(label="**Covariate**",
                     estimate="**Odds**")) |>
  modify_footnote(update = list(estimate ~ NA),
                  abbreviation=TRUE)
t_ym
```

