---
title: 'Lab 6: Conditional Logistic Regression Models'
author: "Benjamin Jarvis"
date: "27 February 2024"
output: pdf_document
fontsize: 11pt
institute: |
  | Institute for Analytical Sociology
  | Linköping University
---

```{r setup , include=F}
knitr::opts_chunk$set(include=T, message=F, warning=F)
options(digits=3)
options(knitr.kable.NA = '')

# For applying colors to text
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
# This is allows us to control text size for Latex output created by R
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

## Objectives

- Reshape data from wide to long format for estimating choice models.

- Estimate and interpret conditional logistic regression models.

- Estimate and interpret decision-maker by alternative-specific covariate interactions.

- Produce in and out of sample predicted probabilities. 

## Reshaping (a.k.a., pivoting) data

- We use pivoting when we want to take data spread across many columns and instead spread that data across rows in a data set, and vice versa.

  - Taking data spread across columns and spreading it across (new) rows is called **reshaping** or **pivoting** to **long format**.

  - Taking data spread across rows and spreading it across (new) columns is called **reshaping** or **pivoting** to **wide format**.
  
- Sometimes this is called "spreading" and "gathering", or "folding" and "unfolding". 
  
- `tidyr` (from `tidyverse`) has the functions `pivot_longer()` and `pivot_wider()` to accomplish this.

- Another popular approach is to use the `melt()` and `dcast()` functions in the `data.table` package.


## Pivoting wide to long^[https://www.garrickadenbuie.com/project/tidyexplain/#spread-and-gather]
  
:::: {.cols data-latex=""}

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{wide} 
:::

::: {.col data-latex="{0.01\textwidth}"}
\vspace{1ex}
:::

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{pivot_longer} 
:::

::: {.col data-latex="{0.01\textwidth}"}
\vspace{1ex}
:::

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{long} 
:::

::::


## Pivoting long to wide^[https://www.garrickadenbuie.com/project/tidyexplain/#spread-and-gather]
  
:::: {.cols data-latex=""}

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{long} 
:::

::: {.col data-latex="{0.01\textwidth}"}
\vspace{1ex}
:::

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{pivot_wider} 
:::

::: {.col data-latex="{0.01\textwidth}"}
\vspace{1ex}
:::

::: {.col data-latex="{0.32\textwidth}"}
\includegraphics[width=\textwidth]{wide} 
:::

::::

## Packages we'll use today

```{r}
library(tidyverse)
library(tidymodels)
library(survival)
library(kableExtra)
library(RColorBrewer)
```

You also should install, but do not need to load, the `Ecdat` package.

## Pivoting wide to long using `tidyr`, 1A

```{r, size='footnotesize'}
relig_income |> slice(1:10)
```

## Pivoting wide to long using `tidyr`, 1B

```{r, size='footnotesize'}
relig_income |> 
  pivot_longer(!religion, 
               names_to = "income", 
               values_to = "count")
```

## Pivoting wide to long using `tidyr`, 2A

```{r, size='footnotesize'}
billboard |> slice(1:7)
```


## Pivoting wide to long using `tidyr`, 2B

```{r , eval=F}
billboard |> 
    pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

## Pivoting wide to long using `tidyr`, 2C

```{r , echo=F, size='footnotesize'}
billboard |> 
    pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

## Pivoting wide to long using `tidyr`, 3A
```{r, size='footnotesize'}
who |> filter(year>2000) |> slice(1:5)
```


## Pivoting wide to long using `tidyr`, 3B

```{r, eval=F}
who %>% select(-iso2,-iso3) |> 
  filter(year>2000) |> 
  pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count"
)
```

## Pivoting wide to long using `tidyr`, 3C

```{r, echo=F, size='footnotesize'}
who %>% select(-iso2,-iso3) |> 
  filter(year>2000) |> 
  pivot_longer(
  cols = new_sp_m014:newrel_f65,
  names_to = c("diagnosis", "gender", "age"),
  names_pattern = "new_?(.*)_(.)(.*)",
  values_to = "count"
)
```


## Revisiting the cars data from the `Ecdat` package

- We were working with data from a survey of California residents from 1996.

- Respondents were solicited to choose a car they would buy from a menu of 6 fictive cars with varying attributes.

- Focus was on assessing the appeal of alternative fuel vehicles.

- Mix of decision-maker level and alternative-level covariates.

## Column names in the `Car` data.

```{r , size="footnotesize"}
Car<-as_tibble(Ecdat::Car)
names(Car)
```

## A view onto some of the `Car` data

```{r , size="scriptsize", echo=F}
Car |> slice(1:16) |> 
  mutate(id=as.integer(row_number())) |> 
  relocate(id,college,hsg2,coml5,choice)
```


## To estimate a `clogit()` model, we need to reshape this data to long format

```{r , size="footnotesize"}
CarLong <- Car |> 
  mutate(id=row_number()) |> # adding an id for respondents
  mutate(choice=str_extract(as.character(choice),"[1-6]"),
    choice=as.integer(choice)) |> # make choice var a #
  pivot_longer(
    cols=!matches(c("id","college","hsg2","coml5","choice")),
    names_pattern="(.+)([1-6])",  # parse colnames w/ regex
    names_to=c(".value","alt")) |> # colnames from "()" above
  mutate(alt=as.integer(alt)) |> 
  relocate(id,college,hsg2,coml5,alt) # rearrange for display
```

## Result of pivot

```{r , size="footnotesize", echo=F}
head(CarLong,n=12)
```

## We still need to make our dependent variable!

```{r}
CarLong <- CarLong %>% 
  mutate(choice=as.integer(choice==alt))
```


## Result, with select columns re-arranged

```{r, size="footnotesize"}
CarLong %>% 
  slice(1:12)
```

## Check the number of responses, alternatives and choices

```{r}
CarLong |> 
  group_by(id) |> 
  summarize(nalts=n(),choices=sum(choice)) |> 
  with(data=_,table(nalts,choices))
```

## Summarizing the data using `gtsummary`

```{r, eval=FALSE, size='small'}
library(gtsummary)
CarLong |> 
  mutate(chof=factor(choice,
                     labels=c("Unchosen","Chosen"))) |> 
  select(chof,type, fuel, price, range) |> 
  tbl_summary(by=chof, type=list(range~"continuous")) |>
  as_gt()
```

## Data summarized using `gtsummary`

```{r, echo=FALSE, size='small'}
library(gtsummary)
CarLong |> 
  mutate(chof=factor(choice,
                     labels=c("Unchosen","Chosen"))) |> 
  select(chof,type, fuel, price, range) |> 
  tbl_summary(by=chof, type=list(range~"continuous")) |>
  as_gt()
```

## A model
```{r, size="footnotesize"}
m1<-clogit(choice~type+fuel+cost+strata(id),data=CarLong)
m1
```

## Presenting model coefficients as odds using `gtsummary`

```{r, size="small", eval=F}
tbl_regression(m1, exponentiate = TRUE) |> 
  modify_header(list(label="**Covariate**",
                     estimate="**Odds**")) |>
  modify_footnote(update = list(estimate ~ NA), 
                  abbreviation=TRUE)
```

## Model coefficients presented using `gtsummary`

```{r, echo=F, size="footnotesize"}
tbl_regression(m1, exponentiate = TRUE) |> 
  modify_header(list(label="**Covariate**",
                     estimate="**Odds**")) |>
  modify_footnote(update = list(estimate ~ NA), 
                  abbreviation=TRUE) |> 
  as_gt()
```



## YOUR TURN

Identify a variable that might explain why methanol or cng vehicles are so much less preferred than gasoline vehicles. 

- Hint: check the correlation between fuel and other variables in the dataset.

Estimate a model where you add in the expected mediator.

Interpret the cng and methanol coefficients before and after you include the mediator.


## SOLUTION

Any suggestions? What did you all find?

## In-Sample Predictions

 - In sample predictions are easy to produce
 - Specify `type="expected"` to get probabilities!
- "Binding" it to the data set is helpful for display

```{r, size="footnotesize"}
CarLong %>% bind_cols(Pr=predict(m1,type="expected")) %>% 
  select(id,alt,choice,type,fuel,cost,Pr)
```

## YOUR TURN

Write some R code that confirms that the probabilities sum to one within values of `id`

## Out of Sample Predictions

- To predict with a new dataset, we need the same covariates used in our model (including `id`!).
- Doesn't even have to be the same size choice set as in the original data.
- Let's see how well our model encapsulates your worst stereotype of American car owners.

```{r, size="footnotesize"}
NewCar<-tibble(id=rep(1,3),
               type=factor(c("sportcar","sportuv","regcar")),
               fuel=factor(c("gasoline","gasoline","electric")),
               cost=c(8,8,1))
```

## Out of Sample Predictions continued

Unfortunately, the `type="expected"` option doesn't seem to work when using new data, so we have some extra work to do using the linear predictions

```{r, size="footnotesize"}
NewCar %>% bind_cols(Za=predict(m1,type="lp",newdata=.)) %>% 
  group_by(id) %>% 
  mutate(Pr=exp(Za)/sum(exp(Za)))
```


## Interactions (1/6)

- What if we want to test the hypothesis that larger households care more about cargo space than smaller households?
 
- We put an appropriate interaction into our model

```{r, size="footnotesize"}
m2<-clogit(choice~type+fuel+cost+space+space*hsg2+strata(id)
           ,data=CarLong)
```

## Interactions (2/6)

```{r, size="footnotesize", echo=F}
m2
```

## Interactions (3/6)

- Note the omitted category for the `hsg2` variable.

- We now have two interpretations to make with the `space` variable.
  - A small household has $e^{0.398}=1.488$ times higher probability to choose a car that has $100\%$ more cargo space than a typical, full size, gas powered car, all else equal.
  - A large household is $e^{0.398-0.448} = 1.488 \times 0.639 =  0.951$ times lower probability to choose a car with $100\%$ more carego space than a typical gas powered car, all else equal.
  
- Actually, we find the opposite of what we expected... but our model is quite poorly specified.

- For interactions with continuous covariates, we might select a few typical values (e.g., 1st quartile, median, and 3rd quartile) of the individual variable and calculate odds for these select values.
 
## Interactions (4/6)

- We could have estimated a model with the interactions specified slightly differently

```{r, size="footnotesize"}
m3<-clogit(choice~type+fuel+cost+space:factor(hsg2)+strata(id)
           ,data=CarLong)
```

## Interactions (5/6)

```{r, echo=F, size="footnotesize"}
m3
```

## Interactions (6/6)

- But notice that these models are actually the same model!

```{r, echo=F, size="footnotesize"}
lapply(list(m2=m2,m3=m3),broom::glance) |> 
  bind_rows(...=_,.id="model") |> 
  select(model,logLik)
```

```{r, echo=F, size="footnotesize"}
bind_cols(pr_m2=predict(m2, type="expected" ),
          pr_m3=predict(m3, type="expected"))
```

## YOUR TURN

Hypothesize and test another decion-maker by alternative interaction. Produce *one* fictive choice set and predictions for that choice set for fictive people who differ in the decision-maker specific attribute you used for your interaction.

## YOUR TURN

Experiment more with `clogit()`. Try:

- Additional covariates
- Predictions based on new choice sets
- Making the "best" model you can.
- Compare predictions between simple models and your "best" model.
- etc.

## Next Up

- Large choice sets
- Linear hypothesis testing
- Model fit
