---
title: "Assignment 1 SNA"
author: "Marc Sparhuber"
format: pdf
editor: source
toc: true
---

## Task 1

> Write an R-script that reads your data. Transform your data to a data frame. Argue about the format you have chosen to analyze large network data. 

```{r}
library(tidyverse)
library(igraph)
library(sna)

# here I load in the data. To not have any issues with how the .txt file is structured I remove everything but the data and add the column names manually

data <- as.data.frame(read_table(file = "C:/Users/marcs/Desktop/Labs/CSS_Labs_early_24/SNA/AssignmentsSNA/Email-Enron.txt",
                   skip = 4,
                   col_names = c("FromNodeID","ToNodeID")))
```

## Task 2

> Make sure that before or while doing any calculations or plots that you handle nonresponse / missing data meaningfully if there is any.

```{r}
# check if there are any NAs

any(is.na(data))

# this indicates that in neither of the two columns any NAs are present
```

## Task 3

> Justify what you do with isolates and multiple components if there are any.

```{r}
# first check how many isolates there are. There are also no multiple components.

isolates(data)

```

## Task 4

> Analyze the density of your network. Create a table that contains further descriptive
network statistics for your network. Please include average degree (in-degree and outdegree), standard deviation of degree (in-degrees and out-degrees), reciprocity, and
transitivity.

```{r}

```

