---
title: "Computer lab 4"
date: last-modified
date-format: "MMMM DD, YYYY"
author: "Richard Öhrvall"
format: pdf
---


# Extra for lecture 4

## Pasting data
One quick way to add data to R, if it's just a small table, is to use the package datapasta. Install the package, copy the table, e.g. in Excel, and paste it into R using data pasta from the addins menu above. 

```{r}


```


## Data transformation
A general advice is to not change the original dataset. Data transformation, e.g. creating new variables etc., can be done in the beginning of an R script or R Markdown file, or in a separate script that is sourced. Or you could create a new analysis dataset that is used in subsequent analyses.

## Tables with summary statistics

```{r}
library(socviz)
library(tidyverse)
library(broom)
library(modelsummary)
library(titanic)

# Get the data
titanic <- titanic_train |> 
  mutate(Pclass = as_factor(Pclass),
         surv = as_factor(Survived))

# You can use * to separate by values of a variable. 1 gives you the whole sample
datasummary(surv * (Age + Fare + Sex) ~ Mean + SD + N + Percent(),
            data = titanic,
            title = "Passengers on the Titanic: Descriptive statistics",
            notes = c("Comments: Data from the Titanic R package."))


# You can also change the output to a data.frame and then append
# that information in a consequtive step, using the add_rows argument
# see https://vincentarelbundock.github.io/modelsummary/articles/datasummary.html#add_rows
datasummary(surv * (Age + Fare + Sex) ~ Mean + SD + N + Percent(),
            data = titanic,
            title = "Passengers on the Titanic: Descriptive statistics",
            notes = c("Comments: Data from the Titanic R package."),
            output = "data.frame")


# If you want the percentages for categorical values to be calculated within
# the groups, you could change the denominator using the denom argument, see below.
datasummary(surv * (Age + Fare + Sex) ~ Mean + SD + N + Percent(denom = Equal(surv)),
            data = titanic,
            title = "Passengers on the Titanic: Descriptive statistics",
            notes = c("Comments: Data from the Titanic R package."))


```

You could also create a data frame with the information you want in the table and the use kable in the knitr package, or kableExtra or some other function, to create a table based on that data frame.



## visualize the results
Let's visualize the results of the logistic regression. We can plot the odds ratios in a graph using ggplot2.

```{r}
## Let's estimate a logistic regression model 

survive_age_log <- glm(Survived ~ Age + Pclass, data = titanic, 
                       family = "binomial"(link = "logit"))

survive_age_log2 <- glm(Survived ~ Age + Sex + Pclass, data = titanic, 
                       family = "binomial"(link = "logit"))

# We set it to -theme_light- (but do try other themes to find someone you like)
theme_set(theme_light())

# Get the results into a data frame (or actually a tibble, i.e. a data frame in Tidyverse)
tidy(survive_age_log, exponentiate = TRUE, conf.int = TRUE)

# The intercept deviates so much from the other odds ratios, and is not that interesting, so we exclude it.
tidy(survive_age_log, exponentiate = TRUE, conf.int = TRUE) |>  
  filter(term != "(Intercept)") |>  
  ggplot(aes(term, estimate)) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
  geom_point() +
  scale_y_continuous(breaks = seq(0, 1.2, by = .2), limits = c(0, 1.2)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high, width = .1))+
  labs(title = "Survival from Titanic. Logistic regression",
       y = "Odds ratios",
       x = "Variables",
       caption = "Data from Titanic R Package")

# If we like to, we can flip the graph to horizontal using -coord_flip-
tidy(survive_age_log, exponentiate = TRUE, conf.int = TRUE) |> 
  filter(term != "(Intercept)") |>  
  mutate(term = case_when(term == "Pclass2" ~ "Passenger class 2",
                          term == "Pclass3" ~ "Passenger class 3",
                          .default = term)) |> 
  ggplot(aes(estimate, term)) +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_point() +
  scale_x_log10(breaks = c(0, 0.1, .5, 1, 1.5)) +
  expand_limits(x = c(0.06, 1.5)) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high, width = .1))+
  labs(title = "Survival from Titanic. Logistic regression",
       x = "Odds ratios",
       y = "Variables",
       caption = "Data from Titanic R Package")


# Combining two models in one graph
# Note the position_dodge in ggplot!
tidy(survive_age_log, exponentiate = TRUE, conf.int = TRUE) |>  
  mutate(model = "A") |> 
  add_row(tidy(survive_age_log2, exponentiate = TRUE, conf.int = TRUE) |> 
            mutate(model = "B")) |> 
  filter(term != "(Intercept)") |>  
  mutate(term = case_when(term == "Pclass2" ~ "Passenger class 2",
                          term == "Pclass3" ~ "Passenger class 3",
                          .default = term)) |> 
  ggplot(aes(estimate, term, color = model)) +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed") +
  geom_point(position = position_dodge(width = .5)) +
  scale_x_log10(breaks = c(0, 0.1, .5, 1, 1.5)) +
  expand_limits(x = c(0.06, 1.5)) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high, width = .1),
                    position = position_dodge(width = .5)) +
  labs(title = "Survival from Titanic. Logistic regression",
       x = "Odds ratios",
       y = "Variables",
       caption = "Data from Titanic R Package")

```


### More about group_by
We have discussed all the verbs used in dplyr, e.g. filter, arrange, mutate, in previous computer labs. Some aspects of group_by are still worth mentioning.

```{r}
# Let's load tidyverse and the gapminder dataset
library(gapminder)

# Let's look at the Gapminder dataset
gapminder

# We can aggregate by group
gapminder |> 
  group_by(continent) |> 
  summarise(avg_lifextp = mean(lifeExp, na.rm = TRUE),
            median_pop = median(pop, na.rm = TRUE))

# But we can also calculate at group level and keep all observations
# without aggregating
gapminder |> 
  group_by(continent) |> 
  mutate(avg_lifextp = mean(lifeExp, na.rm = TRUE),
            median_pop = median(pop, na.rm = TRUE)) 

# You can see that the tibble is still grouped
# From the group_vars function
gapminder |> 
  group_by(continent) |> 
  mutate(avg_lifextp = mean(lifeExp, na.rm = TRUE),
            median_pop = median(pop, na.rm = TRUE)) |> 
  group_vars()

# Remember to ungroup if you want so, 
# otherwise subsequent transformation will be at group level
# You do that using ungroup
gapminder |> 
  group_by(continent) |> 
  mutate(avg_lifextp = mean(lifeExp, na.rm = TRUE),
            median_pop = median(pop, na.rm = TRUE)) |> 
  ungroup()

# If you use summarise, you could also specify 
# that you want the groups to be dropped
gapminder |> 
  group_by(continent, country) |> 
  summarise(avg_lifextp = mean(lifeExp, na.rm = TRUE),
            median_pop = median(pop, na.rm = TRUE),
            .groups = "drop") 

# In order to change multiple columns at once, we can use across
# Here we summarise all columns from lifeExp to gdpPerCap
# we use the new anonymous functions in base R, i.e. \(x)
gapminder |> 
  summarise(across(lifeExp:gdpPercap, \(x) mean(x, na.rm = TRUE)))

# You can, of course, change x to something else
gapminder |> 
  summarise(across(lifeExp:gdpPercap, \(vars) mean(vars, na.rm = TRUE)))

# You could also use the tilde operator, but I recommend \(x)
# gapminder |> 
#   summarise(across(lifeExp:gdpPercap, ~mean(.x, na.rm = TRUE)))


# And here by continent
gapminder |> 
  group_by(continent) |> 
  summarise(across(lifeExp:gdpPercap, \(x) mean(x, na.rm = TRUE)))


# Here all columns that end with p in their name
gapminder |> 
  group_by(continent) |> 
  summarise(across(ends_with("p"), \(x) mean(x, na.rm = TRUE)))



# Or all numeric columns
gapminder |> 
  group_by(continent) |> 
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))

# We can change the column name, using the .names argument
# which uses glue syntax
gapminder |> 
  group_by(continent) |> 
  summarise(across(ends_with("p"), \(x) mean(x, na.rm = TRUE),
            .names = "{col}_mean"))

# we can calculate more than one measurement at once
gapminder |> 
  group_by(continent) |> 
  summarise(across(ends_with("p"), list(medel = mean, sd = sd),
                   .names = "{col}_{fn}"))

# Or we can specify our own function
my_func <- function(x){
  y <- log(x) / x
  return(y)
}

gapminder |> 
  mutate(across(ends_with("p"), \(x) my_func(x)))

```
